### 2015-05-27 papers/histogram/thesis/thesis.tex
minor thesis revisions

We identify and develop efficient numerical methods for determining
thermodynamic properties of the square-well fluid in order to test
square-well density functional theories. The liquid-vapor phase
interface is an interesting regime for testing density functional
theories, but unbiased Monte Carlo simulations are incapable of
sampling the low energy fluid states which dominate the partition
function at low liquid-state temperatures. Previous works have
developed several generic Monte Carlo histogram methods for collecting
statistics on low energy system states, but little or no literature
exists on these methods' systematic comparison, as well as their
application to the square-well fluid. The square-well fluid in
particular introduces application challenges not manifest in
traditional models for testing and benchmarking such numerical
techniques (e.g. the Ising model).

@@

Employing biased Monte Carlo simulations, sketched out in Algorithm
\ref{alg:biased_MC}, allows one to construct weight functions that
favor some region of state space over others, as transitions to states
with higher weights are always accepted, whereas transitions to states
with lower weights may be rejected, artificially preventing the
simulated system from leaving interesting regions of state
space. Crucially, the bias introduced by weights can be reversed when
computing system properties from sampling statistics, as the weight of
a particular state is directly proportional to the probability bias of
that state; that is, a state with a weight of 2 will be sampled twice
as often as it would have been with a weight of 1. If we wish to
convert a histogram $H\p{X}$ of observations in a biased Monte Carlo
simulation into a numerical unnormalized density of states $\tilde
D\p{X}$ in $X$, we therefore divide the histogram by the corresponding
weights $w\p{X}$, i.e.
\begin{align}
  \tilde D\p{X}=\f{H\p{X}}{w\p{X}}.
  \label{eq:dos}
\end{align}
The normalized density of states $D\p{X}$ in $X$ is then
\begin{align}
  D\p{X}=\f{\tilde D\p{X}}{\sum_X\tilde D\p{X}}
  =\f{H\p{X}/w\p{X}}{\sum_XH\p{X}/w\p{X}}.
  \label{eq:dos_norm}
\end{align}
In general, the domain and shape of the weight function will depend on
the desired yields (e.g.  density of states, heat capacity) of a
simulation. A histogram method is simply an algorithm or procedure for
determining a weight function appropriate for a particular simulation.

\begin{algorithm}[tb]
  \caption{Biased Monte Carlo fluid simulation}
  \label{alg:biased_MC}
  \begin{alg}

@@

The most common weight array $w\p{E}$ used by physicists for what is
called a canonical Monte Carlo simulation involves choosing a
particular temperature $T_0$, and using weights proportional to the
Boltzmann factor at that temperature, i.e.
\begin{align}
  w\p{E}=e^{-E/kT_0},
\end{align}
where there is no reason to normalize $w\p{E}$ due to the fact that
only ratios of weights, as per (\ref{eq:move_prob}), are ever used in
simulation. The fact that only ratios of weights are used in
simulation makes $w\p{E}$ scale-invariant.

@@

Having covered the basics of the square-well fluid as well as unbiased
and biased Monte Carlo simulations, we can now discuss several
histogram methods for constructing energy weights. In Section
\ref{sec:flat_histogram}, we will introduce the flat histogram method,
which is not directly applicable to the study of most systems, but
which forms the basis of many other methods. We will then introduce,
in Section \ref{sec:simple_flat}, the ``simple flat'' method, which we
designed to be the simplest viable histogram method, and against which
we can compare more complicated methods. Next, in Sections
\ref{sec:wang_landau}--\ref{sec:optimized_ensemble} we introduce three
published methods: Wang-Landau, transition matrix Monte Carlo, and the
optimized ensemble. The last of these methods we will not actually
implement and analyze because, unlike most histogram methods, it was
designed to optimize weights rather than construct them. We will,
however, borrow ideas from the optimized ensemble to design a
histogram method of our own, a hybrid of the optimized ensemble and
transition matrix Monte Carlo, in Section \ref{sec:oetmmc}. Finally,
we will cover some general implementation details and loose ends in
Section \ref{sec:loose_ends}

@@

Unlike the previous histogram methods, the transition matrix Monte
Carlo method does not use weights, and merely compute the density of
states $D\p{E}$ of a system\cite{tmmc}. The density of states can in
turn be used to determine flat histogram weights via
(\ref{eq:flat_weights}). TMMC introduces a new object: the energy
transition matrix, $T$, whose components $T_{ij}$ are the
probabilities that a system will transition {\it to} a state with
energy $E_i$ {\it from} a given state with energy $E_j$,
i.e. $T_{ij}=P\p{E_j\to E_i}$.

@@

Knowing the transition matrix of a system determines, among other
properties, the system's density of states. Consider an ensemble of
unbiased Monte Carlo simulations with a distribution of energies at a
time $t$ given by $\hat D^{t}\p{E}$, such that the probability of any
given simulation to have an energy $E_i$ at a time $t$ is $\hat
D_i^{t}=\hat D^{t}\p{E_i}$. The probability $\hat D_i^{t}$ can be
expressed in terms of the distribution $\hat D^{t-1}\p{E}$ one time
step (i.e. one Monte Carlo move) prior by
\begin{align}
  \hat D_i^{t}=\sum_jP\p{E_i\to E_j}\hat D_j^{t-1} =\sum_j T_{ij}\hat
  D_j^{t-1}.
  \label{eq:transition_evolution}
\end{align}
At equilibrium, the probability distributions do not vary with time,
meaning $\hat D^{t}=\hat D^{t-1}$. Furthermore, the distribution of
states in an equilibrium ensemble of unbiased Monte Carlo simulations
is precisely the density of states $D\p{E}$, which means
$D_i=\sum_jT_{ij}D_j$.  Expressing this condition for all energies
with a state density vector $D$ and transition matrix $T$,
\begin{align}
  D=TD. \label{eq:dos_eigen}
\end{align}
Finding the density of states of a system can thus be reduced to
finding the eigenvector of the transition matrix which has a
corresponding unit eigenvalue. \fixme{why is this eigenvector unique?}
Once one has determined the density of states of a system, one can
initialize weights via the flat histogram method,
i.e. (\ref{eq:flat_weights}).

@@

All histogram methods introduced in this paper thus far have focused
on determining a weight array $w\p{E}$ for a flat energy histogram
$H\p{E}$, in order to sample and collect statistics on all energies
equally. This motivation considers the quantity of statistics, but not
their quality. Biased simulations collect statistics on low energy
states by first getting into these states, and then rejecting moves
which move the system into higher energy states. Statistics on low
energy states will therefore generally be highly correlated, as they
are based on many samples of only a few low energy regions of a
system's energy landscape. The optimized
ensemble\cite{optimized_ensemble} attempts to address the
autocorrelation between low energy samples by finding weights to
maximize the rate at which a simulation makes round trips between low
and high energy states. Maximizing the round trip rate, in turn,
maximizes the rate at which a simulation makes independent,
uncorrelated samples of low energies.

@@

The algorithm provided in \cite{optimized_ensemble} to construct
weights given by (\ref{eq:optimized_weights}) finds the diffusivity
$\alpha$ by solving for it in
(\ref{eq:walker_current_definition}). This procedure accepts flat
histogram weights, or equivalently the density of states, as an input,
and is not meant to construct weights from scratch. We therefore will
not be comparing the optimized ensemble to other methods in this
paper. We do, however, credit \cite{optimized_ensemble} for motivating
the hybrid OE-TMMC method in Section \ref{sec:oetmmc}.

\subsection{The hybrid OE-TMMC method}
\label{sec:oetmmc}

@@

The hybrid OE-TMMC method thus first runs Algorithm \ref{alg:tmmc} to
collect a transition histogram and compute flat histogram weights,
then computes the diffusivity $\alpha\p{E}$ via
(\ref{eq:<de>})-(\ref{eq:microscopic_diffusitivy}), and finally
computes optimized weights as given by
(\ref{eq:optimized_weights}). Unlike the optimized ensemble, OE-TMMC
does not require any extra simulation routines, and is therefore
straightforward to implement and compare to other histogram methods.

@@

Our analysis will involve comparing errors in system properties
computed from simulation results. As we do not know the true values of
these system properties, we will compare standard simulation results
against those from ``gold standard'' simulations, which run for
significantly longer than the standard $10^5N^2$ iterations. For our
gold standard simulations, we use the TMMC method and initialize until
it collects $10^4$ optimistic samples of the minimum important energy
(i.e. until $s_o\p{E_-}\ge 10^4$). \fixme{why TMMC?} To increase
confidence in low-temperature gold standard results, we initialized
them with $kT_{\t{min}}=0.1\epsilon$. After initialization, the gold
standard runs simulate for $10^{10}$ iterations ($10^{10}N$
moves). While the entire initialization and simulation process takes
around two hours for the largest regular simulations, the largest
golden simulations initialize for $\sim11$ hours, and simulate for a
week. To control for iteration number in the results from any given
set of simulations, we compare against ``converged flat''
simulations. Converged flat simulations read in the transition matrix
from golden simulations to construct a weight array, and simulate for
the same number of iterations as the standard simulations.

@@

The energy histogram in Figure \ref{fig:arrays} shows that TMMC and
OE-TMMC spend considerably less time at important low energies than
the simple flat and vanilla Wang-Landau methods. At face value, this
behavior of TMMC and OE-TMMC is undesirable, as these are the most
interesting energies for low temperatures, and the most difficult
energies to sample.

High energy histogram counts at low energies mean that a simulation
has collected many statistics on these energies, but the energy
histogram contains no information about how correlated these
statistics are. To determine how effective different simulations are
at collecting uncorrelated statistics, we can look at the pessimistic
sample rate $r_p\p{E}$, which is the mean number of simulation
iterations required to sample a given energy $E$. This rate is roughly
equal to the number of iterations $I$ in a simulation and the
pessimistic sample rate $s_p\p{E}$ by $r_p\p{E}=I/s_p\p{E}$; the
greater the number of iterations $I$, the better this
approximation. Figure \ref{fig:sample_rates} shows pessimistic sample
rates averaged over 30 simulations with $N=25$. Lower pessimistic
sample rates mean that a simulation takes less time to collect
statistics which are guaranteed to be uncorrelated. By this metric,
TMMC and OE-TMMC again perform worse than simple flat and vanilla
Wang-Landau.

@@

Histograms and sampling rates let us speculate about the quality of a
simulations, but the most important deliverable of any simulation is
accurate physical results. We therefore compute, via the formula in
(\ref{eq:XT}), the temperature dependence of the square-well fluid's
potential energy
\begin{align}
  U=\bk{E}_T,
  \label{eq:U}
\end{align}
the heat capacity
\begin{align}
  C_V\p{T}=\f{\partial U}{\partial T}
  =\bk{\p{\f{E}{kT}}^2}_T-\bk{\f{E}{kT}}_T^2,
  \label{eq:CV}
\end{align}
and the configurational entropy \fixme{is this correct?}
\begin{align}
  S_{\t{config}}\p{T}=k\bk{\f{E}{kT}+\ln Z}_T
  -k\bk{\f{E}{kT}+\ln Z}_{T\to\infty}.
  \label{eq:S}
\end{align}
All of these properties are directly computable from the density of
states $D\p{E}$, and are shown in Figures
\ref{fig:internal_energy}--\ref{fig:config_entropy} for $N=25$
simulations, averaged over 30 simulations. When looking at these
figures, one should remember that we are only confident in results
down to $kT=0.2\epsilon$ for regular simulations, and $kT=0.1\epsilon$
for golden ones. With the exception of Wang-Landau, which has a heavy
low-energy bias, all simulations agree fairly well on these properties
of the square-well fluid.

Though Figures \ref{fig:internal_energy}--\ref{fig:config_entropy} are
interesting for studying the square-well fluid, they do not help us
resolve the efficacy of different histogram methods. These figures do,
however, show that the square-well fluid has a phase transition around
$kT\approx0.5\epsilon$, which justifies our choice of
$kT_{\t{min}}=0.2\epsilon$ and ensures that the histogram methods are
indeed performing the nontrivial task of exploring the liquid-gas
phase boundary of the square-well fluid. For the purposes of this
paper, we will neglect to prove that this phase transition is indeed
between liquid and gas phases, and take this fact as given. \fixme{say
  more convincingly}

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\textwidth]{figs/internal-energy.pdf}
  \caption[Internal energy]
  {Specific internal energy curves averaged over 30 $N=25$
    simulations.}
  \label{fig:internal_energy}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{figs/heat-capacity.pdf}
  \caption[Heat capacity]
  {Specific heat capacity curves averaged over 30 $N=25$ simulations.}
  \label{fig:heat_capacity}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{figs/config-entropy.pdf}
  \caption[Configurational entropy]
  {Specific configurational entropy curves averaged over 30 $N=25$
    simulations.}
  \label{fig:config_entropy}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{figs/u-err.pdf}
  \caption[Internal energy error scaling]
  {Errors in specific internal energy, averaged over 30 $N=25$
    simulations.}
  \label{fig:internal_energy_err}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{figs/s-comps.pdf}
  \caption[Configurational entropy error comparisons]
  {Specific configurational entropy errors, plotted against those from
    converged flat histogram weight simulations.}
  \label{fig:s_comps}
\end{figure}

Figures \ref{fig:u_comps}--\ref{fig:s_comps} provide a different
visualization of the same information as in Figures
\ref{fig:u_scaling}--\ref{fig:s_scaling}, plotting the errors of all
histogram methods against those of converged flat histogram weight
simulations. The fact that some points fall below the diagonal
reference line is strange, and indicates that some regular simulations
are outperforming the converged flat ones. We do not understand why
such events occur, as converged flat simulations should be using
high-quality golden data to construct weight arrays, and should
therefore outperform all standard simulations. \fixme{comment on why
  some methods might do better than CFW?}

As with the arrays, the error comparison figures demonstrate the
inability of Wang-Landau to sample a system without a predefined
energy range. The simple flat method, vanilla Wang Landau, and TMMC
seem to give comparable errors, except for some anomalous cases with
$N=18$ and $N=21$ in which simple flat has large errors. We do not
know the reason for these anomalies, but they could occur if the
method terminates before finding energies near $E_-$, in which case it
would get stuck in simulation at low energies with canonical weights
for the minimum temperature $kT_{\t{min}}=0.2\epsilon$. Finally, these
error figures further confirm earlier suspicions of OE-TMMC's
underperformance relative to TMMC, despite the fact that it is
supposed to provide an improvement over the TMMC weight array.

@@

After implementing several histogram methods and applying them to
simulation of the $\lambda=1.3$, $\eta=0.3$ square-well fluid, we
found that Wang-Landau, if given a predefined energy range,
consistently yields some of the smallest errors and highest low-energy
sampling rates of all histogram methods, with comparable results from
TMMC and the simple flat method. Without prior knowledge of the
minimum important energy of a system, however, Wang-Landau can get
stuck at low energies, in which case it is incapable of
recovering. Wang-Landau is therefore most appropriate for studying
systems in which the entire energy range is known a priori, or in
which only a given energy range is of particular interest. The simple
flat method, while generally performing well, sometimes fails to
initialize. If gone unnoticed, such failures can be costly in terms of
computation time and resources. Our implementation of the optimized
ensemble using the transition matrix resulted in lower sampling rates
and higher errors than the flat histogram TMMC weights. We suspect
that the optimized ensemble is failing to improve TMMC because the
theory behind it neglects the discretization of energies.

\fixme{make conclusions more satisfying?}

@@

I wish to thank, first and foremost, my advisor Dr. David Roundy, for
mentoring me and providing me a wealth of knowledge and expertise in
various subjects of physics, numerical methods, and programming
nuances. I would also like thank Paho Lurie-Gregg, with whom I
frequently exchange and discuss codes and ideas, and whose hard
polyhedra Monte Carlo simulations I used as the template for my own
codes. Lastly, I would like to thank thank Eva Morgun, who prompted me
to {\it act} on my textbook enthusiasm for physics, and get involved
in the fun and interesting world of scientific research. \fixme{Thank
  Jansen and McIntyre}

>>>>

We identify and develop efficient numerical methods for determining
thermodynamic properties of the square-well fluid in order to test
square-well density functional theories. The liquid-vapor phase
interface is an interesting regime for testing density functional
theories, but unbiased Monte Carlo simulations are incapable of
effectively sampling the low energy fluid states which dominate the
partition function at low liquid-state temperatures. Previous works
have developed several generic Monte Carlo histogram methods for
collecting statistics on low energy system states, but little or no
literature exists on these methods' systematic comparison, as well as
their application to the square-well fluid. The square-well fluid in
particular introduces application challenges not manifest in
traditional models for testing and benchmarking such numerical
techniques (e.g. the Ising model).

@@

This sort of biased simulation is called Metropolis-Hastings Monte
Carlo sampling. Employing Metropolis-Hastings Monte Carlo simulations,
sketched out in Algorithm \ref{alg:biased_MC}, allows one to construct
weight functions that favor some region of state space over others, as
transitions to states with higher weights are always accepted, whereas
transitions to states with lower weights may be rejected, artificially
preventing the simulated system from leaving interesting regions of
state space. Crucially, the bias introduced by weights can be reversed
when computing system properties from sampling statistics, as the
weight of a particular state is directly proportional to the
probability bias of that state; that is, a state with a weight of 2
will be sampled twice as often as it would have been with a weight of
1. If we wish to convert a histogram $H\p{X}$ of observations in a
Metropolis-Hastings Monte Carlo simulation into a numerical
unnormalized density of states $\tilde D\p{X}$ in $X$, we therefore
divide the histogram by the corresponding weights $w\p{X}$, i.e.
\begin{align}
  \tilde D\p{X}=\f{H\p{X}}{w\p{X}}.
  \label{eq:dos}
\end{align}
The normalized density of states $D\p{X}$ in $X$ is then
\begin{align}
  D\p{X}=\f{\tilde D\p{X}}{\sum_{X'}\tilde D\p{X'}}
  =\f{H\p{X}/w\p{X}}{\sum_{X'}H\p{X'}/w\p{X'}}.
  \label{eq:dos_norm}
\end{align}
In general, the domain and shape of the weight function will depend on
the desired yields (e.g.  density of states, heat capacity) of a
simulation. A histogram method is simply an algorithm or procedure for
determining a weight function appropriate for a particular simulation.

\begin{algorithm}[tb]
  \caption{Metropolis-Hastings Monte Carlo fluid simulation}
  \label{alg:biased_MC}
  \begin{alg}

@@

The most common weight array $w\p{E}$ used by physicists for what is
called a canonical (or Metropolis) Monte Carlo simulation involves
choosing a particular temperature $T_0$, and using weights
proportional to the Boltzmann factor at that temperature, i.e.
\begin{align}
  w\p{E}=e^{-E/kT_0},
\end{align}
where there is no reason to normalize $w\p{E}$ due to the fact that
only ratios of weights, as per (\ref{eq:move_prob}), are ever used in
simulation. The fact that only ratios of weights are used in
simulation makes $w\p{E}$ scale-invariant.

@@

Having covered the basics of the square-well fluid as well as Monte
Carlo fluid simulation, we can now discuss several histogram methods
for constructing energy weights. In Section \ref{sec:flat_histogram},
we will introduce the flat histogram method, which is not directly
applicable to the study of most systems, but which forms the basis of
many other methods. We will then introduce, in Section
\ref{sec:simple_flat}, the ``simple flat'' method, which we designed
to be the simplest viable histogram method, and against which we can
compare more complicated methods. Next, in Sections
\ref{sec:wang_landau}--\ref{sec:optimized_ensemble} we introduce three
published methods: Wang-Landau, transition matrix Monte Carlo, and the
optimized ensemble. The last of these methods we will not actually
implement and analyze because, unlike most histogram methods, it was
designed to optimize pre-existing weights rather than construct
them. We will, however, borrow ideas from the optimized ensemble to
design a histogram method of our own, a hybrid of the optimized
ensemble and transition matrix Monte Carlo, in Section
\ref{sec:oetmmc}. Finally, we will cover some general implementation
details and loose ends in Section \ref{sec:loose_ends}

@@

Unlike the previous histogram methods, the transition matrix Monte
Carlo method does not use weights, and merely computes the density of
states $D\p{E}$ of a system\cite{tmmc}. The density of states can in
turn be used to determine flat histogram weights via
(\ref{eq:flat_weights}). TMMC introduces a new object: the energy
transition matrix, $T$, whose components $T_{ij}$ are the
probabilities that a system will transition {\it to} a state with
energy $E_i$ {\it from} a given state with energy $E_j$,
i.e. $T_{ij}=P\p{E_j\to E_i}$.

@@

Knowing the transition matrix of a system determines, among other
properties, the system's density of states. Consider an ensemble of
unbiased Monte Carlo simulations with a distribution of energies at a
time $t$ given by $\hat D^{t}\p{E}$, such that the probability of any
given simulation to have an energy $E_i$ at a time $t$ is $\hat
D_i^{t}=\hat D^{t}\p{E_i}$. The probability $\hat D_i^{t}$ can be
expressed in terms of the distribution $\hat D^{t-1}\p{E}$ one time
step (i.e. one Monte Carlo move) prior by
\begin{align}
  \hat D_i^{t}=\sum_jP\p{E_i\to E_j}\hat D_j^{t-1} =\sum_j T_{ij}\hat
  D_j^{t-1}.
  \label{eq:transition_evolution}
\end{align}
At equilibrium, the probability distributions do not vary with time,
meaning $\hat D^{t}=\hat D^{t-1}$. Furthermore, the distribution of
states in an equilibrium ensemble of unbiased Monte Carlo simulations
is precisely the density of states $D\p{E}$, which means
$D_i=\sum_jT_{ij}D_j$.  Expressing this condition for all energies
with a state density vector $D$ and transition matrix $T$,
\begin{align}
  D=TD. \label{eq:dos_eigen}
\end{align}
Finding the density of states of a system can thus be reduced to
finding the eigenvector of the transition matrix which has a
corresponding unit eigenvalue. Once one has determined the density of
states of a system, one can initialize weights via the flat histogram
method, i.e. (\ref{eq:flat_weights}).

@@

All histogram methods introduced in this paper thus far have focused
on determining a weight array $w\p{E}$ for a flat energy histogram
$H\p{E}$, in order to sample and collect statistics on all energies
equally. This motivation considers the quantity of statistics, but not
their quality. Metropolis-Hastings Monte Carlo simulations collect
statistics on low energy states by first getting into these states,
and then rejecting moves which move the system into higher energy
states. Statistics on low energy states will therefore generally be
highly correlated, as they are based on many samples of only a few low
energy regions of a system's energy landscape. The optimized
ensemble\cite{optimized_ensemble} attempts to address the
autocorrelation between low energy samples by finding weights to
maximize the rate at which a simulation makes round trips between low
and high energy states. Maximizing the round trip rate, in turn,
maximizes the rate at which a simulation makes independent,
uncorrelated samples of low energies.

@@

The algorithm provided in \cite{optimized_ensemble} to construct
weights given by (\ref{eq:optimized_weights}) finds the diffusivity
$\alpha$ by solving for it in
(\ref{eq:walker_current_definition}). This procedure accepts flat
histogram weights, or equivalently the density of states, as an input,
and is not meant to construct weights from scratch. We therefore will
not be comparing the optimized ensemble to other methods in this
paper. We do, however, credit \cite{optimized_ensemble} for motivating
the hybrid DOTMMC method in Section \ref{sec:oetmmc}.

\subsection{The diffusion-optimized transition matrix Monte Carlo
  (DOTMMC) method}
\label{sec:oetmmc}

@@

The diffusion-optimized transition matrix Monte Carlo (DOTMMC) method,
a hybrid of TMMC and the optimized ensemble, thus first runs Algorithm
\ref{alg:tmmc} to collect a transition histogram and compute flat
histogram weights, then computes the diffusivity $\alpha\p{E}$ via
(\ref{eq:<de>})-(\ref{eq:microscopic_diffusitivy}), and finally
computes optimized weights as given by
(\ref{eq:optimized_weights}). Unlike the optimized ensemble, DOTMMC
does not require any extra simulation routines, and is therefore
straightforward to implement and compare to other histogram methods.

@@

Our analysis will involve comparing errors in system properties
computed from simulation results. As we do not know the true values of
these system properties, we will compare standard simulation results
against those from ``gold standard'' simulations, which run for
significantly longer than the standard $10^5N^2$ iterations. For our
gold standard simulations, we use the TMMC method and initialize until
it collects $10^4$ optimistic samples of the minimum important energy
(i.e. until $s_o\p{E_-}\ge 10^4$). To increase confidence in
low-temperature gold standard results, we initialized them with
$kT_{\t{min}}=0.1\epsilon$. After initialization, the gold standard
runs simulate for $10^{10}$ iterations ($10^{10}N$ moves). While the
entire initialization and simulation process takes around two hours
for the largest regular simulations, the largest golden simulations
initialize for $\sim11$ hours, and simulate for a week. To control for
iteration number in the results from any given set of simulations, we
compare against ``converged flat'' simulations. Converged flat
simulations read in the transition matrix from golden simulations to
construct a weight array, and simulate for the same number of
iterations as the standard simulations.

@@

The energy histogram in Figure \ref{fig:arrays} shows that TMMC and
DOTMMC spend considerably less time at important low energies than
the simple flat and vanilla Wang-Landau methods. At face value, this
behavior of TMMC and DOTMMC is undesirable, as these are the most
interesting energies for low temperatures, and the most difficult
energies to sample.

High energy histogram counts at low energies mean that a simulation
has collected many statistics on these energies, but the energy
histogram contains no information about how correlated these
statistics are. To determine how effective different simulations are
at collecting uncorrelated statistics, we can look at the pessimistic
sample rate $r_p\p{E}$, which is the mean number of simulation
iterations required to sample a given energy $E$. This rate is roughly
equal to the number of iterations $I$ in a simulation and the
pessimistic sample rate $s_p\p{E}$ by $r_p\p{E}=I/s_p\p{E}$; the
greater the number of iterations $I$, the better this
approximation. Figure \ref{fig:sample_rates} shows pessimistic sample
rates averaged over 30 simulations with $N=25$. Lower pessimistic
sample rates mean that a simulation takes less time to collect
statistics which are guaranteed to be uncorrelated. By this metric,
TMMC and DOTMMC again perform worse than simple flat and vanilla
Wang-Landau.

@@

Histograms and sampling rates let us speculate about the quality of a
simulations, but the most important deliverable of any simulation is
accurate physical results. We therefore compute, via the formula in
(\ref{eq:XT}), the temperature dependence of the square-well fluid's
potential energy
\begin{align}
  U=\bk{E}_T,
  \label{eq:U}
\end{align}
the heat capacity
\begin{align}
  C_V\p{T}=\f{\partial U}{\partial T}
  =\bk{\p{\f{E}{kT}}^2}_T-\bk{\f{E}{kT}}_T^2,
  \label{eq:CV}
\end{align}
and the configurational entropy
\begin{align}
  S_{\t{config}}\p{T}=S\p{T}-S\p{T\to\infty}
  =k\bk{\f{E}{kT}+\ln Z}_T-k\bk{\f{E}{kT}+\ln Z}_{T\to\infty}.
  \label{eq:S}
\end{align}
All of these properties are directly computable from the density of
states $D\p{E}$, and are shown in Figures
\ref{fig:internal_energy}--\ref{fig:config_entropy} for $N=25$
simulations, averaged over 30 simulations. When looking at these
figures, one should remember that we are only confident in results
down to $kT=0.2\epsilon$ for regular simulations, and $kT=0.1\epsilon$
for golden ones. With the exception of Wang-Landau, which has a heavy
low-energy bias, all simulations agree fairly well on these properties
of the square-well fluid.

Though Figures \ref{fig:internal_energy}--\ref{fig:config_entropy} are
interesting for studying the square-well fluid, they do not help us
resolve the efficacy of different histogram methods. These figures do,
however, show that the square-well fluid has a phase transition around
$kT\approx0.5\epsilon$, which justifies our choice of
$kT_{\t{min}}=0.2\epsilon$ and ensures that the histogram methods are
indeed performing the nontrivial task of exploring the liquid-gas
phase boundary of the square-well fluid. For the purposes of this
paper, we will neglect to prove that this phase transition is indeed
between liquid and gas phases, and take this fact as given.

\begin{figure}[p]
  \centering
  \includegraphics[width=0.75\textwidth]{figs/internal-energy.pdf}
  \caption[Internal energy]
  {Specific internal energy curves averaged over 30 $N=25$
    simulations.}
  \label{fig:internal_energy}
\end{figure}

\begin{figure}[p]
  \centering
  \includegraphics[width=0.75\textwidth]{figs/heat-capacity.pdf}
  \caption[Heat capacity]
  {Specific heat capacity curves averaged over 30 $N=25$ simulations.}
  \label{fig:heat_capacity}
\end{figure}

\begin{figure}[p]
  \centering
  \includegraphics[width=0.75\textwidth]{figs/config-entropy.pdf}
  \caption[Configurational entropy]
  {Specific configurational entropy curves averaged over 30 $N=25$
    simulations.}
  \label{fig:config_entropy}
\end{figure}

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\textwidth]{figs/u-err.pdf}
  \caption[Internal energy error scaling]
  {Errors in specific internal energy, averaged over 30 $N=25$
    simulations.}
  \label{fig:internal_energy_err}
\end{figure}

@@

\begin{figure}[p]
  \centering
  \includegraphics[width=0.7\textwidth]{figs/s-comps.pdf}
  \caption[Configurational entropy error comparisons]
  {Specific configurational entropy errors, plotted against those from
    converged flat histogram weight simulations.}
  \label{fig:s_comps}
\end{figure}

Figures \ref{fig:u_comps}--\ref{fig:s_comps} provide a different
visualization of the same information as in Figures
\ref{fig:u_scaling}--\ref{fig:s_scaling}, plotting the errors of all
histogram methods against those of converged flat histogram weight
simulations. The fact that some points fall below the diagonal
reference line is strange, and indicates that some regular simulations
are outperforming the converged flat ones. We do not understand why
such events occur, as converged flat simulations should be using
high-quality golden data to construct weight arrays, and should
therefore outperform all standard simulations.

As with the arrays, the error comparison figures demonstrate the
inability of Wang-Landau to sample a system without a predefined
energy range. The simple flat method, vanilla Wang Landau, and TMMC
seem to give comparable errors, except for some anomalous cases with
$N=18$ and $N=21$ in which simple flat has large errors. We do not
know the reason for these anomalies, but they could occur if the
method terminates before finding energies near $E_-$, in which case it
would get stuck in simulation at low energies with canonical weights
for the minimum temperature $kT_{\t{min}}=0.2\epsilon$. Finally, these
error figures further confirm earlier suspicions of DOTMMC's
underperformance relative to TMMC, despite the fact that it is
supposed to provide an improvement over the TMMC weight array.

@@

After implementing several histogram methods and applying them to
simulation of the $\lambda=1.3$, $\eta=0.3$ square-well fluid, we
found that Wang-Landau, if given a predefined energy range,
consistently yields some of the smallest errors and highest low-energy
sampling rates of all histogram methods, with comparable results from
TMMC and the simple flat method. Without prior knowledge of the
minimum important energy of a system, however, Wang-Landau can get
stuck at low energies, in which case it is incapable of
recovering. Wang-Landau is therefore most appropriate for studying
systems in which the entire energy range is known a priori, or in
which only a given energy range is of particular interest. The simple
flat method, while generally performing well, sometimes fails to
initialize. If gone unnoticed, such failures can be costly in terms of
computation time and resources. Our implementation of the optimized
ensemble using the transition matrix, i.e. DOTMMC, resulted in lower
sampling rates and higher errors than the flat histogram TMMC
weights. Though we have not completely ruled out that our
implementation of DOTMMC contains bugs, we suspect that the optimized
ensemble is failing to improve TMMC either because
\begin{enumerate*}[label=\roman*)]
\item the theory behind the optimized ensemble neglects the
  discretization of energies, or
\item DOTMMC is highly sensitive to sampling error in the transition
  matrix.
\end{enumerate*}
The first of these suspicions can be tested by simulating larger
systems with a denser specific energy spectrum, and for which
discretization should become asymptotically negligible; the second
suspicion can be tested simply by initializing via TMMC for longer,
thereby reducing errors in the transition matrix, before computing
diffusion-optimized weights.

@@

I wish to thank, first and foremost my advisor, David Roundy, for
mentoring me and providing me a wealth of knowledge and expertise in
various subjects of physics, numerical methods, and programming
nuances. I would also like thank Paho Lurie-Gregg, with whom I
frequently exchange and discuss codes and ideas, and whose hard
polyhedra Monte Carlo simulations I used as the template for my own
square-well codes. Lastly, I would like to thank thank Eva Morgun, who
prompted me to {\it act} on my textbook enthusiasm for physics, and
get involved in the fun and interesting world of scientific
research. \fixme{Thank Jansen and McIntyre}
